### 2014-11-11 doc/thesis/typedlua.tex
Updating chapter 3 to include the integer type

Typed Lua is an optional type system for Lua, and its main goal is to
allow programmers to combine static and dynamic typing in Lua.
In practice, Typed Lua extends the syntax of Lua to introduce
optional type annotations and class-based object-oriented
programming through the definition of classes, interfaces, and
modules.
In addition, Typed Lua performs local type inference \citep{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

@@

Lua values can have one of eight tags:
\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
Typed Lua includes types for the first six.
Typed Lua also includes a syntactical extension that programmers can use
to define the types of \emph{userdata}, and uses it to define the type of \emph{thread}. 
In this section, we present the Typed Lua types that may appear on annotations.
We explain all Typed Lua types and syntactical extensions along this chapter.

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} & ::= \; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Number} \; | \;
  \textit{String}\\
\textit{basetype} & ::= \; \textbf{boolean} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettupletype}\\
\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettupletype}\\
\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettupletype} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
\textit{keytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{value} \; | \;
  \textbf{any}\\
\textit{rettype} & ::= \; \textit{type} \; | \; \textit{rettupletype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

@@

Typed Lua uses subtyping to order types.
Any first-level type is a subtype of \texttt{value}.
The rest of the subtyping relationship is standard:
union types are supertypes of their parts,
\texttt{boolean}, \texttt{number}, and \texttt{string} are supertypes
of their respective literal types,
function types are related by contravariance on the input
and covariance on the output, table types have width subtyping,
with depth subtyping on \texttt{const} fields, tuple and vararg types
are covariant.

@@

In the following example, we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}
    local function factorial (n:number):number
      if n == 0 then
        return 1
      else
        return n * factorial(n - 1)
      end
    end
    local x = 5
    print(factorial(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type \texttt{number} to \texttt{x}, and this example compiles
without any errors.
The compiler does not use the literal type \texttt{5},
instead of the base type \texttt{number}, because this would prevent
programmers from assigning other numeric values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

Typed Lua allows programmers to combine statically typed code with
dynamically typed code, as we show in the following example:
\begin{verbatim}
    local function absolute (n:number):number
      if n < 0 then
        return -n
      else
        return n
      end
    end

@@

The function \texttt{distance} receives two parameters of type \texttt{any}
and returns a value of type \texttt{number}.
The compiler assigns the dynamic type \texttt{any} to the input
parameters of \texttt{distance} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
Even though we did not annotated the return type of \texttt{distance},
the compiler is able to infer its return type because it is local
and not recursive.

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.
More precisely, we can call \texttt{absolute} inside \texttt{distance}
because the expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{number}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not a number, making the dynamically typed code crash the
statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

@@

To match the semantics of Lua function calls, the \emph{default} operation
mode of Typed Lua adds a variadic tail to the parts of a function
type whenever the programmer does not specify one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(number, value*) -> (number, nil*)},
and the type of \texttt{distance} is \texttt{(any, any, value*) -> (number, nil*)}.
If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
them, as the type signature lets \texttt{absolute} receive any number of
extra arguments.
If we call \texttt{absolute} with fewer arguments, Typed Lua uses \texttt{nil} to
replace the missing arguments and in the case of \texttt{absolute} generates
an error, as \texttt{nil} is neither a subtype of \texttt{number} nor consistent
with \texttt{number}.

Typed Lua also provides a \emph{strict} operation mode, and programmers can
use this mode if they want to catch arity mismatch of function calls.
In the strict mode, Typed Lua does not add a variadic tail to the
parts of a function type unless the programmer explicitly declares one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(number) -> (number)},
and the type of \texttt{distance} is \texttt{(any, any) -> (number)}.
If we call \texttt{absolute} with either extra or missing arguments,
Typed Lua generates errors about arity mismatch.

A variadic type can only appear in the tail position of a tuple,
because Lua takes only the first value of any expression that appears
in an expression list that is not in tail position.
The following example shows the interaction between multiple returns
and expression lists:
\begin{verbatim}
    local function multiple ()
      return 2, "foo"
    end
    local function sum (x:number, y:number)
      return x + y
    end
    local x, y, z = multiple(), multiple()
    print(sum(multiple(), multiple()))
\end{verbatim}

In the default mode, the type of \texttt{multiple} is
\texttt{(value*) -> (number, string, nil*)},
and the type of \texttt{sum} is \texttt{(number, number, value*) -> (number, nil*)}.
In the right hand side of the multiple assignment, only the first
value produced by the first call to multiple gets used,
so the type of the right hand side is
\texttt{(number, number, string, nil*)},
and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
are respectively \texttt{number}, \texttt{number}, and \texttt{string}.
The same tuple is assigned to the call to \texttt{sum} and it
compiles without errors, as the first two components of the tuple
are consistent with the types of the parameters, and the other
components are consistent with \texttt{value}.

In the strict mode, the type of \texttt{multiple} is \texttt{() -> (number, string)},
and the type of \texttt{sum} is \texttt{(number, number) -> (number)}.
Like in the default mode, only the first value produced by the
first call to multiple gets used in the right hand side of the
multiple assignment;
and the type of the right hand side is also
\texttt{(number, number, string, nil*)},
because the strict mode affects only the type of function declarations
and multiple assignments that appear in function calls.
Unlike in the default mode, the type of the call to \texttt{sum} is
\texttt{(number, number, string)}, and does not include the
variadic tail to catch arity mismatch.
Thus, this example compiles with error because the call to \texttt{sum}
passes an unexpected extra argument of type \texttt{string}.

@@

Lua programmers often overload the input parameters of functions,
and use the \texttt{type} function to inspect the tag of the
input parameters to take different actions depending on what
those tags are.
The simplest case overloads on just a single parameter:
\begin{verbatim}
    local function overload (s1:string, s2:string|number)
      if type(s2) == "string" then
        return s1 .. s2
      else
        -- string.rep : (string, number, string) -> (string)
        return string.rep(s1, s2)
      end
    end
\end{verbatim}

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and usually returns \texttt{nil} as the first value plus an
error message or other data that describes the error
when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:number, divisor:number):
          (number, number)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) / divisor
        return q, r
      end 
    end
\end{verbatim}

Typed Lua also includes a syntactic sugar for this idiom:
we can annotate the return type of \texttt{idiv} with
\texttt{(number, number)?} to denote the same union.
The parentheses are always necessary in this case, because
\texttt{t?} is syntactic sugar for \texttt{t|nil},
while \texttt{(t)?} is syntactic sugar for \texttt{(t)|(nil, string)}.

@@

When Typed Lua finds a union of tuples in the right hand side of
a declaration, it stores this union in a special type environment
with a fresh name and assigns \emph{projection types} to the
variables that appear unannotated in the left hand side of the declaration.
In our example, Typed Lua assigns the union of tuples
\texttt{(number, number, nil*)|(nil, string, nil*)} to a type variable $X$.
Thus, variable \texttt{q} gets the type $X_{1}$
that is equivalent to \texttt{number|nil},
and variable \texttt{r} gets the type $X_{2}$
that is equivalent to \texttt{number|string}.

Projection types do not appear in the subtyping and consistent-subtyping
relations because Typed Lua actually replaces them with the union of the
corresponding component of each tuple.
If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates against all tuples in the union.
In our example, $X$ is \texttt{(number, number, nil*)|(nil, string, nil*)}
outside of the \texttt{if} statement,
but \texttt{(number, number, nil*)} inside the \texttt{then} block,
and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
Thus, variable \texttt{q} has type \texttt{number} and variable \texttt{r} also
has type \texttt{number} inside the \texttt{then} block;
and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
has type \texttt{string} inside the \texttt{else} block.

@@

Although Typed Lua has just one framework to represent tables,
it has three different syntaxes for defining the type of a table:
\begin{align}
\label{syn:table1}
\texttt{`\{'} \; \textit{keytype} \; \texttt{`:'} \; \textit{type} \; \texttt{`\}'}\\
\label{syn:table2}
\texttt{`\{'} \; \textit{type} \; \texttt{`\}'}\\
\label{syn:table3}
\texttt{`\{'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type} \;
\{\texttt{`,'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}\} \;
[\texttt{`,'} \; \textit{type}] \; \texttt{`\}'}
\end{align}
The syntax \ref{syn:table1} defines table types for maps:
it is written \texttt{\{k:t\}},
and maps to the table type \texttt{\{k:t|nil\}}.
This table type represents a map that maps values of type
\texttt{k} to values of type \texttt{t|nil},
as Lua returns \texttt{nil} when we use a non-existing key to index a table.
The syntax \ref{syn:table2} defines table types for arrays:
it is written \texttt{\{t\}},
and maps to the table type \texttt{\{number:t|nil\}}.
This table type is just a syntactic sugar to the definition of a
map that maps values of type \texttt{number} to values of type \texttt{t|nil}.
The syntax \ref{syn:table3} defines table types for records with an
optional array part: it is written
$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
and maps to the same table type.
This table type represents a record that maps the literal types
$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.

Next, we show one example of table type to type a map from strings to numbers:
\begin{verbatim}
    local t:{string:number} = { foo = 1 } 
    local x:number = t.foo               -- compile-time error
    local y:number? = t.bar              -- y gets nil
    local z:number = t["bar"] or 0       -- z gets 0 
\end{verbatim}

The second line of this example gives a type error because we are
attempting to assign a value of type \texttt{number|nil} to a
variable that accepts only values of type \texttt{number}.
Although the field \texttt{bar} does not exist in \texttt{t}, the third
line of this example does not give a type error because the
annotated type matches the type of the values that can be stored in
\texttt{t}.
The last line shows that the \texttt{or} idiom is also useful to
give a default value to a missing table field.

@@

In this example, the type of both variables \texttt{x} and \texttt{y}
is \texttt{string|nil}.
An inconvenience aspect of making the types of maps and arrays always
include the type \texttt{nil} is to overload the programmers,
as they need to use \texttt{or} or \texttt{if} to narrow the type of
the elements they are accessing.

@@

We can also use the style of the inferred type to define heterogeneous
tuples:
\begin{verbatim}
    local album:{1:string, 2:number, 3:string} =
      { "Transformer", 1972, "Lou Reed" }
\end{verbatim}

@@

\begin{verbatim}
    local interface Element
      info:number
      next:Element?
    end
\end{verbatim}

@@

Now we can use \texttt{Element} to annotate a function that
inserts an element at the beginning of a list:
\begin{verbatim}
    local function insert (e:Element?, v:number):Element 
      return { info = v, next = e }
    end
\end{verbatim}

@@

Typed Lua uses table types along with the self type to
represent objects, as we show in the following example:
\begin{verbatim}
    local Shape = { x = 0, y = 0 }
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

@@

Typed Lua combines this mechanism along with interfaces to allow
programmers to define classes in a similar fashion to the definition
of modules, as we show in the following example:
\begin{verbatim}
    interface Shape
      x, y:number
      const new:(number, number) => (self)
      const move:(number, number) => ()
    end
    local Shape = { x = 0, y = 0 }
    const function Shape:new (x:number, y:number)
      local s = setmetatable({}, { __index = self })
      s.x = x
      s.y = y
      return s
    end
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    return Shape
\end{verbatim}

@@

\begin{verbatim}
    local Shape = require "shape"
    interface Circle
      x, y, radius:number
      const new:(number, number, value) => (self)
      const move:(number, number) => ()
      const area:() => (number)
    end
    local Circle = setmetatable({}, { __index = Shape })
    Circle.radius = 0
    const function Circle:new (x:number, y:number, radius:value)
      local c = setmetatable(Shape:new(x, y), { __index = self })
      c.radius = tonumber(radius) or 0
      return c
    end
    const function Circle:area ()
      return math.pi * self.radius * self.radius
    end
    return Circle
\end{verbatim}

>>>>

Typed Lua is an optional type system for Lua, and its main goal is to
allow programmers to combine static and dynamic typing in Lua.
In practice, Typed Lua extends the syntax of Lua to introduce
optional type annotations.
In addition, Typed Lua performs local type inference \citep{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

@@

Lua values can have one of eight tags:
\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
Typed Lua includes types for the first six.
Typed Lua also includes a syntactical extension that programmers can use
to define the types of \emph{userdata}, and uses it to define the type \emph{thread}. 
In this section, we present the Typed Lua types that may appear on annotations.
We explain all Typed Lua types and syntactical extensions along this chapter.

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} & ::= \; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} & ::= \; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettupletype}\\
\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettupletype}\\
\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettupletype} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
\textit{keytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{value} \; | \;
  \textbf{any}\\
\textit{rettype} & ::= \; \textit{type} \; | \; \textit{rettupletype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

@@

Typed Lua uses subtyping to order types.
Any first-level type is a subtype of \texttt{value}.
The rest of the subtyping relationship is standard:
union types are supertypes of their parts,
\texttt{boolean}, \texttt{integer}, \texttt{number}, and \texttt{string}
are supertypes of their respective literal types,
\texttt{integer} is subtype of \texttt{number},
function types are related by contravariance on the input
and covariance on the output, table types have width subtyping,
with depth subtyping on \texttt{const} fields, tuple and vararg types
are covariant.

@@

In the following example, we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}
    local function factorial (n:integer):integer
      if n == 0 then
        return 1
      else
        return n * factorial(n - 1)
      end
    end
    local x = 5
    print(factorial(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type \texttt{integer} to \texttt{x}, and this example compiles
without any errors.
The compiler does not use the literal type \texttt{5},
instead of the base type \texttt{integer}, because this would prevent
programmers from assigning other integer values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

Typed Lua allows programmers to combine statically typed code with
dynamically typed code, as we show in the following example:
\begin{verbatim}
    local function absolute (n:integer):integer
      if n < 0 then
        return -n
      else
        return n
      end
    end

@@

The function \texttt{distance} receives two parameters of type \texttt{any}
and returns a value of type \texttt{integer}.
The compiler assigns the dynamic type \texttt{any} to the input
parameters of \texttt{distance} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
Even though we did not annotated the return type of \texttt{distance},
the compiler is able to infer its return type because it is local
and not recursive.

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not an integer number,
because in the semantics of Lua the minus operator can result in a
value that is not an integer number.
More precisely, we can call \texttt{absolute} inside \texttt{distance}
because the expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{integer}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not an integer, making the dynamically typed code crash the
statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

@@

To match the semantics of Lua function calls, the \emph{default} operation
mode of Typed Lua adds a variadic tail to the parts of a function
type whenever the programmer does not specify one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(integer, value*) -> (integer, nil*)},
and the type of \texttt{distance} is \texttt{(any, any, value*) -> (integer, nil*)}.
If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
them, as the type signature lets \texttt{absolute} receive any number of
extra arguments.
If we call \texttt{absolute} with fewer arguments, Typed Lua uses \texttt{nil} to
replace the missing arguments and in the case of \texttt{absolute} generates
an error, as \texttt{nil} is neither a subtype of \texttt{integer} nor consistent
with \texttt{integer}.

Typed Lua also provides a \emph{strict} operation mode, and programmers can
use this mode if they want to catch arity mismatch of function calls.
In the strict mode, Typed Lua does not add a variadic tail to the
parts of a function type unless the programmer explicitly declares one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(integer) -> (integer)},
and the type of \texttt{distance} is \texttt{(any, any) -> (integer)}.
If we call \texttt{absolute} with either extra or missing arguments,
Typed Lua generates errors about arity mismatch.

A variadic type can only appear in the tail position of a tuple,
because Lua takes only the first value of any expression that appears
in an expression list that is not in tail position.
The following example shows the interaction between multiple returns
and expression lists:
\begin{verbatim}
    local function multiple ()
      return 2, "foo"
    end
    local function sum (x:integer, y:integer)
      return x + y
    end
    local x, y, z = multiple(), multiple()
    print(sum(multiple(), multiple()))
\end{verbatim}

In the default mode, the type of \texttt{multiple} is
\texttt{(value*) -> (integer, string, nil*)},
and the type of \texttt{sum} is \texttt{(integer, integer, value*) -> (integer, nil*)}.
In the right hand side of the multiple assignment, only the first
value produced by the first call to multiple gets used,
so the type of the right hand side is
\texttt{(integer, integer, string, nil*)},
and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
are respectively \texttt{integer}, \texttt{integer}, and \texttt{string}.
The same tuple is assigned to the call to \texttt{sum} and it
compiles without errors, as the first two components of the tuple
are consistent with the types of the parameters, and the other
components are consistent with \texttt{value}.

In the strict mode, the type of \texttt{multiple} is \texttt{() -> (integer, string)},
and the type of \texttt{sum} is \texttt{(integer, integer) -> (integer)}.
Like in the default mode, only the first value produced by the
first call to multiple gets used in the right hand side of the
multiple assignment;
and the type of the right hand side is also
\texttt{(integer, integer, string, nil*)},
because the strict mode affects only the type of function declarations
and multiple assignments that appear in function calls.
Unlike in the default mode, the type of the call to \texttt{sum} is
\texttt{(integer, integer, string)}, and does not include the
variadic tail to catch arity mismatch.
Thus, this example compiles with error because the call to \texttt{sum}
passes an unexpected extra argument of type \texttt{string}.

@@

Lua programmers often overload the input parameters of functions,
and use the \texttt{type} function to inspect the tag of the
input parameters to take different actions depending on what
those tags are.
The simplest case overloads on just a single parameter:
\begin{verbatim}
    local function overload (s1:string, s2:string|integer)
      if type(s2) == "string" then
        return s1 .. s2
      else
        -- string.rep : (string, integer, string) -> (string)
        return string.rep(s1, s2)
      end
    end
\end{verbatim}

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and usually returns \texttt{nil} as the first value plus an
error message or other data that describes the error
when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:integer, divisor:integer):
          (integer, integer)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) / divisor
        return q, r
      end 
    end
\end{verbatim}

Typed Lua also includes a syntactic sugar for this idiom:
we can annotate the return type of \texttt{idiv} with
\texttt{(integer, integer)?} to denote the same union.
The parentheses are always necessary in this case, because
\texttt{t?} is syntactic sugar for \texttt{t|nil},
while \texttt{(t)?} is syntactic sugar for \texttt{(t)|(nil, string)}.

@@

When Typed Lua finds a union of tuples in the right hand side of
a declaration, it stores this union in a special type environment
with a fresh name and assigns \emph{projection types} to the
variables that appear unannotated in the left hand side of the declaration.
In our example, Typed Lua assigns the union of tuples
\texttt{(integer, integer, nil*)|(nil, string, nil*)} to a type variable $X$.
Thus, variable \texttt{q} gets the type $X_{1}$
that is equivalent to the type \texttt{integer|nil},
and variable \texttt{r} gets the type $X_{2}$
that is equivalent to the type \texttt{integer|string}.

Projection types do not appear in the subtyping and consistent-subtyping
relations because Typed Lua actually replaces them with the union of the
corresponding component of each tuple.
If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates against all tuples in the union.
In our example, $X$ is \texttt{(integer, integer, nil*)|(nil, string, nil*)}
outside of the \texttt{if} statement,
but \texttt{(integer, integer, nil*)} inside the \texttt{then} block,
and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
Thus, variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
has type \texttt{integer} inside the \texttt{then} block;
and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
has type \texttt{string} inside the \texttt{else} block.

@@

Although Typed Lua has just one framework to represent tables,
it has three different syntaxes for defining the type of a table:
\begin{align}
\label{syn:table1}
\texttt{`\{'} \; \textit{keytype} \; \texttt{`:'} \; \textit{type} \; \texttt{`\}'}\\
\label{syn:table2}
\texttt{`\{'} \; \textit{type} \; \texttt{`\}'}\\
\label{syn:table3}
\texttt{`\{'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type} \;
\{\texttt{`,'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}\} \;
[\texttt{`,'} \; \textit{type}] \; \texttt{`\}'}
\end{align}
The syntax \ref{syn:table1} defines table types for maps:
it is written \texttt{\{k:t\}},
and maps to the table type \texttt{\{k:t|nil\}}.
This table type represents a map that maps values of type
\texttt{k} to values of type \texttt{t|nil},
as Lua returns \texttt{nil} when we use a non-existing key to index a table.
The syntax \ref{syn:table2} defines table types for arrays:
it is written \texttt{\{t\}},
and maps to the table type \texttt{\{integer:t|nil\}}.
This table type is just a syntactic sugar to the definition of a
map that maps values of type \texttt{integer} to values of type \texttt{t|nil}.
The syntax \ref{syn:table3} defines table types for records with an
optional array part: it is written
$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
and maps to the same table type.
This table type represents a record that maps the literal types
$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.

Next, we show one example of table type to type a map from strings to integers:
\begin{verbatim}
    local t:{string:integer} = { foo = 1 } 
    local x:integer = t.foo               -- compile-time error
    local y:integer? = t.bar              -- y gets nil
    local z:integer = t["bar"] or 0       -- z gets 0 
\end{verbatim}

The second line of this example gives a type error because we are
attempting to assign a value of type \texttt{integer|nil} to a
variable that accepts only values of type \texttt{integer}.
Although the field \texttt{bar} does not exist in \texttt{t}, the third
line of this example does not give a type error because the
annotated type matches the type of the values that can be stored in
\texttt{t}.
The last line shows that the \texttt{or} idiom is also useful to
give a default value to a missing table field.

@@

In this example, the type of both variables \texttt{x} and \texttt{y}
is \texttt{string|nil}.
An inconvenient aspect of making the types of maps and arrays always
include the type \texttt{nil} is to overload the programmers,
as they need to use \texttt{or} or \texttt{if} to narrow the type of
the elements they are accessing.

@@

We can also use the style of the inferred type to define heterogeneous
tuples:
\begin{verbatim}
    local album:{1:string, 2:integer, 3:string} =
      { "Transformer", 1972, "Lou Reed" }
\end{verbatim}

@@

\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

@@

Now we can use \texttt{Element} to annotate a function that
inserts an element at the beginning of a list:
\begin{verbatim}
    local function insert (e:Element?, v:integer):Element 
      return { info = v, next = e }
    end
\end{verbatim}

@@

Typed Lua uses table types along with the self type to
represent objects, as we show in the following example:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

@@

Typed Lua combines this mechanism along with interfaces to allow
programmers to define classes in a similar fashion to the definition
of modules, as we show in the following example:
\begin{verbatim}
    interface Shape
      x, y:number
      const new:(number, number) => (self)
      const move:(number, number) => ()
    end
    local Shape = { x = 0.0, y = 0.0 }
    const function Shape:new (x:number, y:number)
      local s = setmetatable({}, { __index = self })
      s.x = x
      s.y = y
      return s
    end
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    return Shape
\end{verbatim}

@@

\begin{verbatim}
    local Shape = require "shape"
    interface Circle
      x, y, radius:number
      const new:(number, number, value) => (self)
      const move:(number, number) => ()
      const area:() => (number)
    end
    local Circle = setmetatable({}, { __index = Shape })
    Circle.radius = 0.0
    const function Circle:new (x:number, y:number, radius:value)
      local c = setmetatable(Shape:new(x, y), { __index = self })
      c.radius = tonumber(radius) or 0.0
      return c
    end
    const function Circle:area ()
      return math.pi * self.radius * self.radius
    end
    return Circle
\end{verbatim}
